---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 11,
  fig.height = 6.5,
  dpi = 120
)
```

# chaidr

<!-- badges: start -->
<!-- badges: end -->

chaidr is a base R implementation of the CHAID (Chi-squared Automatic
Interaction Detection) and Exhaustive CHAID decision tree algorithms, as
specified in the IBM SPSS Statistics Algorithms documentation (Kass 1980;
Biggs, de Ville, and Suen 1991). The core fitting and prediction routines
depend only on base R -- visualization backends and 'partykit' integration
are optional soft dependencies.

日本語版 README（README.ja.md）はソースリポジトリに同梱されています。

## Features

* Standard **CHAID** and **Exhaustive CHAID** tree growing with multi-way
  splits.
* Predictors: nominal, ordinal (with a *floating* category for missing
  values), and continuous (automatically discretized into quantile bins).
* Responses: nominal (Pearson chi-squared test), ordinal (Goodman
  row-effects model), and continuous (one-way ANOVA F test).
* Frequency weights via `freq`, reproducing results on pre-aggregated data.
* Prediction of classes, class probabilities, and node membership with
  `predict()`.
* Model inspection: node tables (`chaid_table()`), decision rules
  (`chaid_rules()`), and predictor importance (`chaid_importance()`).
* Evaluation: gains and lift analysis (`chaid_gains()`) and validation on
  holdout data (`chaid_validate()`).
* Visualization: base graphics `plot()`, 'Graphviz' DOT export
  (`chaid_dot()`, `chaid_graphviz()`), interactive 'plotly' trees
  (`chaid_plotly()`), and conversion to 'partykit' objects
  (`chaid_as_party()`).

## Installation

Once on CRAN, install the released version with:

``` r
install.packages("chaidr")
```

Until then, you can install the development version from a local checkout:

``` r
# from the package root directory
# install.packages("devtools")
devtools::install()
```

## Quick start

The `penguins` data set (bundled with R >= 4.5) contains a mix of
continuous predictors and missing values, both of which CHAID handles
natively.

```{r quickstart}
library(chaidr)

data(penguins)
fit <- chaid(species ~ ., data = penguins,
             control = chaid_control(min_parent = 30, min_child = 10))
print(fit)
```

Continuous predictors such as `flipper_len` are discretized into quantile
bins before growing, and statistically similar bins are merged back
together, yielding multi-way splits. Groups containing `<NA>` show where
missing values were merged as a floating category. Terminal nodes are
marked with `*`.

The same tree can be plotted with base graphics (bars show the class
distribution within each node):

```{r plot}
plot(fit, main = "CHAID: penguins (species)")
```

Prediction uses the standard S3 `predict()` interface:

```{r predict}
pred <- predict(fit, penguins)          # class labels (factor)
mean(pred == penguins$species)          # training accuracy

round(predict(fit, head(penguins, 3), type = "prob"), 3)
```

Decision rules for every terminal node:

```{r rules}
chaid_rules(fit)
```

## Key functions

| Task                 | Functions                                              |
|----------------------|--------------------------------------------------------|
| Model fitting        | `chaid()`, `chaid_control()`                           |
| Prediction           | `predict()`                                            |
| Inspection           | `print()`, `summary()`, `chaid_table()`, `chaid_rules()`, `chaid_importance()` |
| Evaluation           | `chaid_gains()`, `chaid_validate()`                    |
| Visualization        | `plot()`, `chaid_dot()`, `chaid_graphviz()`, `chaid_plotly()` |
| partykit integration | `chaid_as_party()`                                     |

## Documentation

* `vignette("chaidr")` -- a compact introduction (in English).
* `vignette("chaidr-ja")` -- a detailed tutorial covering the algorithm,
  all option settings, evaluation, and visualization (in Japanese).

## References

* Kass, G. V. (1980). An exploratory technique for investigating large
  quantities of categorical data. *Applied Statistics*, 29(2), 119--127.
* Biggs, D., de Ville, B., & Suen, E. (1991). A method of choosing
  multiway partitions for classification and decision trees. *Journal of
  Applied Statistics*, 18(1), 49--62.
* IBM Corp. *IBM SPSS Statistics Algorithms* -- "CHAID and Exhaustive
  CHAID Algorithms".

## License

MIT (c) Osamu Morimoto
