---
title: "Introduction to chaidr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to chaidr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 5,
  out.width = "100%"
)
has_partykit <- requireNamespace("partykit", quietly = TRUE)
```

chaidr implements the CHAID (Chi-squared Automatic Interaction Detection)
and Exhaustive CHAID decision tree algorithms in base R, following the
IBM SPSS Statistics Algorithms specification. Unlike binary trees such as
CART, CHAID produces **multi-way splits** driven by hypothesis tests:
categories of each predictor are merged while they are statistically
similar, and the node is split by the predictor with the smallest
Bonferroni-adjusted p value.

This vignette is a compact tour of the package. A much more detailed
tutorial (algorithm internals, Exhaustive CHAID trade-offs, weights,
misclassification costs, ordinal responses, and every visualization
backend) is available in Japanese in the companion vignette
`vignette("chaidr-ja")`.

## Quick start

The `penguins` data set (bundled with R >= 4.5) mixes continuous
predictors with missing values, both of which CHAID handles natively.

```{r quickstart}
library(chaidr)

data(penguins)
fit <- chaid(species ~ ., data = penguins,
             control = chaid_control(min_parent = 30, min_child = 10))
print(fit)
```

Continuous predictors such as `flipper_len` are discretized into up to 10
quantile bins before growing; statistically similar bins are then merged
back, producing multi-way splits. Groups containing `<NA>` show where
missing values were merged as a *floating* category. Terminal nodes are
marked with `*`.

```{r plot, fig.height=6}
plot(fit, main = "CHAID: penguins (species)")
```

Prediction uses the standard S3 interface:

```{r predict}
pred <- predict(fit, penguins)          # class labels (factor)
mean(pred == penguins$species)          # training accuracy

round(predict(fit, head(penguins, 3), type = "prob"), 3)  # probabilities
predict(fit, head(penguins, 3), type = "node")            # terminal node ids
```

## The algorithm in brief

At each node CHAID repeats three phases:

1. **Merge** -- for every predictor, iteratively merge the pair of
   (allowable) categories with the least significant difference in the
   response, until all remaining pairs are significant
   (`alpha_merge`). Ordinal predictors may merge adjacent categories
   only; nominal predictors may merge any pair.
2. **Split** -- compute a Bonferroni-adjusted p value for each merged
   predictor and split by the best one if adjusted p <= `alpha_split`.
3. **Stop** -- otherwise (or when depth/size limits are reached) the node
   becomes terminal.

The significance test is selected by the response type:

| Response          | Test                                  |
|-------------------|---------------------------------------|
| `factor` (nominal)| Pearson chi-squared (or likelihood ratio G²) |
| `ordered`         | Goodman row-effects model (likelihood ratio H²) |
| `numeric`         | One-way ANOVA F                       |

The main knobs of `chaid_control()` (defaults match the SPSS UI):

| Parameter     | Default | Meaning                                        |
|---------------|---------|------------------------------------------------|
| `alpha_merge` | 0.05    | keep merging while pairwise p exceeds this     |
| `alpha_split` | 0.05    | split when adjusted p is at or below this      |
| `max_depth`   | 3       | maximum tree depth                             |
| `min_parent`  | 100     | minimum cases to attempt a split               |
| `min_child`   | 50      | minimum cases per child node                   |
| `n_bins`      | 10      | target quantile bins for continuous predictors |

## Exhaustive CHAID

`method = "exhaustive"` keeps merging each predictor all the way down to
two categories and then picks the most significant configuration from the
whole merge history, avoiding the local optima of the standard greedy
merge. With the Titanic data it discovers additional structure below a
coarser first split:

```{r titanic}
tit <- as.data.frame(Titanic)
fit_std <- chaid(Survived ~ Class + Sex + Age, data = tit, freq = tit$Freq)
fit_ex  <- chaid(Survived ~ Class + Sex + Age, data = tit, freq = tit$Freq,
                 method = "exhaustive")
print(fit_ex)
```

Note the `freq` argument: frequency weights make the fit on aggregated
data identical to the fit on the expanded case-level data.

The two methods redistribute the Bonferroni penalty differently, which
matters when nominal predictors with many levels are present -- see the
Japanese vignette for benchmarks, a simulation of the false-positive
bias, and practical guidance on choosing between them.

## Reporting

Terminal-node summaries, decision rules (as text, SQL, or R
expressions), and a p-value-based predictor importance:

```{r reporting}
tb <- chaid_table(fit, target = "Gentoo")
tb[, setdiff(names(tb), "rule")]

head(chaid_rules(fit, format = "sql"), 3)

chaid_importance(fit)
```

Gains and lift analysis (which nodes capture the target class most
efficiently), and validation of a fitted tree on holdout data:

```{r gains, fig.height=4.5}
g <- chaid_gains(fit_std, target = "Yes")
print(g)
plot(g)
```

```{r validate}
set.seed(9)
idx <- sample(nrow(penguins), 244)
fit_tr <- chaid(species ~ ., data = penguins[idx, ],
                control = chaid_control(min_parent = 30, min_child = 10))
chaid_validate(fit_tr, penguins[-idx, ])
```

## Visualization

The built-in `plot()` method (shown above) needs no extra packages.
Three optional backends are available:

```{r viz, eval = FALSE}
chaid_graphviz(fit)              # 'Graphviz' via DiagrammeR (publication quality)
chaid_dot(fit, file = "tree.gv") # raw DOT export for the dot CLI
chaid_plotly(fit)                # interactive htmlwidget with hover details
```

A fitted tree can also be converted to a `partykit::party` object, which
opens up the partykit and ggparty plotting ecosystems:

```{r partykit, eval = has_partykit, fig.height=6}
pt <- chaid_as_party(fit, penguins)   # pass the data used for fitting
plot(pt)
```

## Further reading

* `vignette("chaidr-ja")` -- detailed tutorial (in Japanese): algorithm
  internals, standard vs. Exhaustive CHAID, weights, missing-value
  handling, multiplicity adjustment across predictors, comparison with
  CART, misclassification costs, ordinal responses, and all
  visualization backends.

## References

* Kass, G. V. (1980). An exploratory technique for investigating large
  quantities of categorical data. *Applied Statistics*, 29(2), 119--127.
* Biggs, D., de Ville, B., & Suen, E. (1991). A method of choosing
  multiway partitions for classification and decision trees. *Journal of
  Applied Statistics*, 18(1), 49--62.
* IBM Corp. *IBM SPSS Statistics Algorithms* -- "CHAID and Exhaustive
  CHAID Algorithms".
